---
title: "TXT 1 Chapter 2 Exercise"
output: html_notebook
---

############################################################################################ 

# Week 3 Part 02 TXT 1 Ch2 Assignment

# The purpose of this project is to use BRFSS data to create a table of appropriate descriptive statistics and describe the characteristics of transgender participants in the 2014 BRFSS

# Matt Mueller

# 2/7/2024

Questions:

1)  Open the transgender_hc_ch2.csv 2014 BRFSS data file. – 0.5 points

2)  Select the data including only transgender participants. (Hint: filter(TRNSGNDR \< 4)) – 1 point

3)  Select the data including only the variables of interest (Hint: select(TRNSGNDR, PHYSHLTH, X_INCOMG, X_EDUCAG, X_AGEG5YR, X_AGE80, X_RACE)). –1 point

4)  Check the data types of all the variables and fix any that seem incorrect. (Hint: PHYSHLTH and X_AGE80 are numeric; the rest are factors) – 1 point

5)  Based on the BRFSS codebook, code missing values and add category labels appropriately – 1 point.

6)  Choose and conduct appropriate descriptive statistics for all variables in the small data set (Achievements 2 and 3). – 0.5 points

7)  Develop a well-formatted results table, including all variables in the small data set (Achievement 4). – 0.5 points

8)  Add a prolog and comments to your code. – 0.5 points

9)  Summarize the characteristics of transgender survey participants in the 2014 BRFSS. – 1 point

############################################################################################# 

```{r}
# load in the libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tableone)
```

```{r}
# load in the data
df <- read.csv("C:/Users/muell/Downloads/Week3_Part02_TXT1_Ch2_Assignment/Week3_Part02_TXT1_Ch2_Assignment/transgender_hc_ch2.csv")
```

```{r}
# filter the df for rows with TRNSGNDR value less than 4
df_filtered <- filter(df, TRNSGNDR < 4)

rm(df)
```

```{r}
# select these columns from the df
df_selected <- select(df_filtered, TRNSGNDR, PHYSHLTH, 
X_INCOMG, X_EDUCAG, X_AGEG5YR, X_AGE80, X_RACE)

rm(df_filtered)
```

```{r}
# check the datatype of the variables
class(df_selected$TRNSGNDR)
class(df_selected$X_AGE5YR)
class(df_selected$X_RACE)
class(df_selected$X_INCOMG)
class(df_selected$X_EDUCAG)
class(df_selected$X_AGE80)
class(df_selected$PHYSHEALTH)
```

```{r}
# assign the correct datatype to each variable
df_selected$TRNSGNDR <- as.factor(df_selected$TRNSGNDR)
df_selected$X_AGE5YR <- as.factor(df_selected$X_AGEG5YR)
df_selected$X_RACE <- as.factor(df_selected$X_RACE)
df_selected$X_INCOMG <- as.factor(df_selected$X_INCOMG)
df_selected$X_EDUCAG <- as.factor(df_selected$X_EDUCAG)
df_selected$X_AGE80 <- as.numeric(df_selected$X_AGE80)
df_selected$PHYSHLTH <- as.numeric(df_selected$PHYSHLTH)
```

```{r}
# add proper category labels for TRNSGNDR column
df_selected$TRNSGNDR <- factor(df_selected$TRNSGNDR,
                               levels = c(1, 2, 3),
                               labels = c("Transgender male-to-female",
                                          "Transgender female-to-male",
                                          "Transgender gender nonconforming")
                               )

# Change PHYSHLTH 77 and 99 to NA
df_selected <- df_selected %>%
  mutate(PHYSHLTH = na_if(PHYSHLTH, 77)) %>%
  mutate(PHYSHLTH = na_if(PHYSHLTH, 99))

# recode PHYSHLTH 88 to 0
df_selected$PHYSHLTH[df_selected$PHYSHLTH == 88] <- 0

# add category labels to X_INCOMG
df_selected$X_INCOMG <- factor(df_selected$X_INCOMG,
                               levels = c(1, 2, 3, 4, 5, 9),
                               labels = c("Less than $15,000",
                                          "$15,000 to less than $25,000",
                                          "$25,000 to less than $35,000",
                                          "$35,000 to less than $50,000",
                                          "$50,000 or more",
                                          "NA_character_")
                               )

# add category labels to X_EDUCAG
df_selected$X_EDUCAG <- factor(df_selected$X_EDUCAG,
                               levels = c(1, 2, 3, 4, 9),
                               labels = c("Did not graduate High School",
                                          "Graduated High Schoool",
                                          "Attended College or Technical School",
                                          "Graduated College or Technical School",
                                          "NA_character_")
                               )

# add category labels to X_AGEG5YR
df_selected$X_AGE5YR <- factor(df_selected$X_AGEG5YR,
                               levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14),
                               labels = c("Age 18 to 24",
                                          "Age 25 to 29",
                                          "Age 30 to 34",
                                          "Age 35 to 39",
                                          "Age 40 to 44",
                                          "Age 45 to 49",
                                          "Age 50 to 54",
                                          "Age 55 to 59",
                                          "Age 60 to 64",
                                          "Age 65 to 69",
                                          "Age 70 to 74",
                                          "Age 75 to 79",
                                          "Age 80 or older",
                                          "NA_character_")
                               )

# add category labels to X_RACE
df_selected$X_RACE <- factor(df_selected$X_RACE,
                               levels = c(1, 2, 3, 4, 5, 6, 7, 8),
                               labels = c("Black only, non-Hispanic",
                                          "American Indian or Alaskan Native only, Non-Hispanic",
                                          "Asian only, non-Hispanic",
                                          "Native Hawaiian or other Pacific Islander only, Non-Hispanic",
                                          "Other race only, non-Hispanic",
                                          "Multiracial, non-Hispanic",
                                          "Hispanic",
                                          "NA_character_")
                               )

```

```{r}
# check the df for labels
str(df_selected)
```

```{r}
# get the descriptive statistics for physhlth
df_selected %>%
  drop_na(PHYSHLTH) %>%
  summarize(mean_days = mean(PHYSHLTH, na.rm=TRUE),
            med_days = median(PHYSHLTH, na.rm=TRUE),
            IQR_days = IQR(PHYSHLTH),
            sd_days = sd(PHYSHLTH),
            mode_days = names(x=sort(table(PHYSHLTH),decreasing=TRUE))[1])
```

```{r}
# use tableone to create a table
contVars <- c("TRNSGNDR", "X_INCOMG", "X_EDUCAG", "X_AGE5YR", "X_RACE")
catVars <- c("PHYSHLTH", "X_AGE80")

table1 <- CreateTableOne(vars = c(catVars, contVars), data=df_selected, factorVars = catVars)

print(table1, nonnormal = contVars)
```

```{r}
# Over half (58.9%) of the participants show good physical health over the last 30 days reporting 0 days of feeling physically unhealthy in the last 30 days. The age of the participants ranged from 18 to 80 with the most frequent age being between 60-64. The majority of the participants (72.1%) are in the Black only, non-Hispanic category with the American Indian or Alaskan Native only category coming in second at 10.5%. The most common income group among the participants is the $50,000 or more category, but 60.2% of the transgender participants earn under $50,000 per year. The transition status category indicates the most transitions are male-to-female at 52.5% and female-to-male transitions at 30.7%. Gender nonconforming transitions accounted for 16.8% of the participants, suggesting this transition is less common among the participants. 
```
