---
title: "TXT1_Chapter9_Assignment"
output: html_notebook
---

#############################################################################
# TXT1 Chapter 9 Assignment
# The purpose of this assignment is to implement and evaluate linear models using the dist_ssp_amfar_ch9 dataset. We create two linear models, with the second including the metro variable, to determine which variable can be used as a predictor for the distance variable. Through the interpretation of test statistics and comparing the two models, we can determine which variable is a statistically significant predictor. 
# Matt Mueller
# 04/17/2025
# Data: dist_ssp_amfar_ch9.csv in the assignment file
#############################################################################

```{r}
#load the libraries
library(ggplot2)
library(lmtest)
```

## Question 1

```{r}
# load the dataset to variable ssp.data

ssp.data <- read.csv("dist_ssp_amfar_ch9.csv")
```

## Question 2

```{r}
# create model using lm() with cube root of outcome variable dist_SSP and opiod_RxRate as the predictor variable

dist.by.rx <- lm(I(dist_SSP^(1/3)) ~ opioid_RxRate, data=ssp.data)
```


## Question 3

```{r}
# Check the model assumptions

# Step 1: Independent Observations
str(ssp.data)
```

```{r}
# Step 2: Continuous Outcome
is.numeric(ssp.data$dist_SSP)
```

```{r}
# Step 3: Linearity
ggplot(ssp.data, aes(x = opioid_RxRate, y = I(dist_SSP^(1/3)))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  labs(
    x = "Prescriptions per 100 people",
    y = "Cube root of dist to SSP"
  )
```

```{r}
# Step 4: Constant variance
bptest(dist.by.rx)
```

```{r}
# Step 5: Independence of residuals 
dwtest(dist.by.rx)
```

```{r}
# Step 6: Normality of residuals
residuals <- data.frame(res = residuals(dist.by.rx))

ggplot(residuals, aes(x = res)) +
  geom_histogram(bins = 20) +
  theme_bw() +
  labs(
    x = "Residuals",
    y = "Count",
    title = "Histogram of Residuals"
  )

```

## Question 4:

```{r}
# interpret the model results using summary()
summary(dist.by.rx)
```

### Interpretation : Only around 0.22% of the variability in the cube-root distance is explained by the opioid prescription rate. The p-value of the F-test is 0.2968, so the model is not significant which indicates that we fail to reject the null hypothesis that counties with higher opioid prescription rates are farther away. 

## Question 5

```{r}
# Create model using lm() with cube root of outcome variable dist_SPP and opioid_RxRate and metro as the predictor variables

dist.by.rx.met <- lm(I(dist_SSP^(1/3)) ~ opioid_RxRate + metro, data=ssp.data)
```

## Question 6

```{r}
# use summary() and confint() and interpret these results

summary(dist.by.rx.met)

```
### Interpretation: The p-value of the F-statistic is 5.78e-11, which indicates a significant test result. The results show that non-metro counties have a higher distance than metro counties even when opioid prescription rate is not a significant predictor. 


```{r}
confint(dist.by.rx.met)
```


### Interpretation: The effect of the metro variable is positive indicating non-metro counties have a higher distance.


## Question 7

```{r}
# compare the larger and smaller models using the partial-F test

# use anova() with dist.by.rx and dist.by.rx.met

anova(dist.by.rx, dist.by.rx.met)
```

## Question 8

### Interpretation of the Partial-F test: The partial-F test is significant with a p-value of 1.174e-11. This indicates that adding the metro variable helps explain a signficant amount of variance in the cube-root distance compared to opioid prescription rate on its own. Therefore, the metro variable is a statistically significant predictor of the distance variable.  
